---
title: "Untitled"
author: "Scott Jackson"
date: "January 17, 2017"
output: html_document
---

# Review

```{r}
# couple of tricks for merging columns

# 1. Check column names, and/or alter `by.x = , by.y = `

# 2. Be vewwy vewwy careful about merging factors
#    - easy to mess up
#    - only really works when factors are the same, with the same levels
#    - ALTERNATIVE: merge as strings
#      - use as.character() to convert to strings
#      - use identical(unique(x)) to look for discrepencies, or unique(x) %in% unique(y)

load("student_data/Mortality.RData")
load("student_data/Population.RData")

ls()

sum(colnames(Mortality) %in% colnames(Population))
which(colnames(Mortality) %in% colnames(Population))
length(which(colnames(Mortality) %in% colnames(Population)))

colnames(Population) %in% colnames(Mortality)

colnames(Population)[colnames(Population) %in% colnames(Mortality)]

unique(Population$Country) %in% unique(Mortality$Country)
unique(Population$Country)[!unique(Population$Country) %in% unique(Mortality$Country)]
unique(Mortality$Country)[!unique(Mortality$Country) %in% unique(Population$Country)]

unique(Population$Admin1) %in% unique(Mortality$Admin1)
unique(Mortality$Admin1) %in% unique(Population$Admin1)

unique(Population$SubDiv) %in% unique(Mortality$SubDiv)
unique(Mortality$SubDiv) %in% unique(Population$SubDiv)

unique(Population$Year) %in% unique(Mortality$Year)
unique(Mortality$Year) %in% unique(Population$Year)

unique(Population$Sex) %in% unique(Mortality$Sex)
unique(Mortality$Sex) %in% unique(Population$Sex)

unique(Population$Frmat) %in% unique(Mortality$Frmat)
unique(Mortality$Frmat) %in% unique(Population$Frmat)

mort.merge1 <- merge(Mortality, Population) # all = FALSE
mort.merge2 <- merge(Mortality, Population, all = TRUE)

head(mort.merge1)
nrow(Population)
nrow(mort.merge1)
nrow(mort.merge2)
nrow(Mortality)

```

# Graphics overview

http://accidental-art.tumblr.com/

## The importance of graphing

```{r}
?anscombe
plot(anscombe)
example("anscombe")
```


## Two systems

  - "base" graphics
  - "grid" graphics
  - See *R Graphics* by Paul Murrell for a very thorough "deep-dive" into these implementations:  https://www.stat.auckland.ac.nz/~paul/RG2e/

### Base graphics
  - more "primitive"
  - can also be much faster than grid
  - lots of isolated plot functions
  - lots of weird little parameters to worry about
  - some handy quick-plot functions from base graphics:
    - `plot()` : a generic function that will often produce something interesting, can do scatterplots with a formula `y ~ x`
    - `hist()` : histogram of a vector
    - `boxplot()` : boxplot, using a formula `variable ~ grouping.factor`

### Grid graphics
  - better for more "high-level" plots
  - much more complex underlying system (IMHO)
  - more powerful and extendable
  - the basis of extremely popular packages such as `lattice` and `ggplot2`

## Graphics packages

  - many packages will have some kind of graphics functions in them, for visualizing specific kinds of things
  - these will often be in base graphics, "pre-canned", etc.
  - the two best more general-purpose graphics packages are `lattice` and `ggplot2`
  - `ggplot2` is another Hadley Wickham project, and has the most documentation, examples, and is currently most widely used

# Using `ggplot2`

## A "Grammar of Graphics"
  - based on work by Leland Wilkinson, a "theory" of graphic display of data
  - Hadley implemented this framework as his dissertation
  - `ggplot2` is the overhauled version of the original implementation, and it is still under active development
  - the idea is to implement a general system for mapping data variables to *aesthetics* (Wilkinson's term), i.e., visual dimensions (x, y, color, etc.)
  - allows for high-level control of the structure of a plot, to allow for flexible, powerful visualizations
  - the place for documentation: http://docs.ggplot2.org/current/
  - another good site: http://www.cookbook-r.com/Graphs/
  
## Basic structure and concepts

  - unlike base graphics, the *value* of `ggplot2` objects is a complex object, and `print()`-ing them renders the graph.
  - the value of the object is built up by adding *layers*
  - start with a call to `ggplot()` that specifies the data and some mappings of variables to aesthetics
  - add "geoms" to actually render something
  - add other things for fine control over graphical parameters, etc.

## Graphing the `iris` package

```{r}
library(ggplot2)
myris <- iris
summary(myris)

ggplot(data = myris, aes(x = Sepal.Length)) + geom_histogram()

ggplot(myris, aes(x = Sepal.Length)) + geom_histogram(binwidth = .5)

ggplot(myris, aes(x = Sepal.Length)) + geom_histogram(binwidth = .01)

hist(myris$Sepal.Length)

mybaseplot <- ggplot(myris, aes(x = Sepal.Length))

mybaseplot + geom_histogram(binwidth = .2)

names(mybaseplot)
str(mybaseplot)

base.density <- mybaseplot + geom_density(adjust = .5, fill = "grey")
base.density

mydensity <- geom_density(adjust = .5, fill = "grey")

mybaseplot + mydensity

```

```{r}
colors()
?RGB
```

### Saving plots

```{r}
pdf("myplot.pdf", width = 10, height = 4)
ggplot(myris, aes(x = Sepal.Length)) + geom_histogram() + 
  facet_grid(Species ~ .)
dev.off()

?Devices

# look into other PDF readers if you want to be 
# able to leave the file "open" (i.e., displayed)
# on your computer while you are editing/changing
# the file
```

## Some more parameters

```{r}
pdf("bunch of plots.pdf")
ggplot(myris, aes(x = Sepal.Length, y = Sepal.Width)) + geom_point()

ggplot(myris, aes(x = Sepal.Length, y = Sepal.Width)) + geom_point() + theme_classic()

myscatterplot <- ggplot(myris, aes(x = Sepal.Length, y = Sepal.Width)) + geom_point()
myscatterplot
myscatterplot + theme_classic()
myscatterplot + theme_minimal()

ggplot(myris, aes(x = Sepal.Length, y = Sepal.Width)) + 
  geom_point(aes(color = Species), size = 10) + 
  theme_minimal()

ggplot(myris, aes(x = Sepal.Length, y = Sepal.Width)) + 
  geom_point(aes(color = Petal.Width), size = 10) + 
  theme_minimal()

ggplot(myris, aes(x = Sepal.Length, y = Sepal.Width)) + 
  geom_point(aes(color = Species, size = Petal.Width)) + 
  theme_minimal()

ggplot(myris, aes(x = Sepal.Length, y = Sepal.Width)) + 
  geom_point() + 
  geom_smooth()

ggplot(myris, aes(x = Sepal.Length, y = Sepal.Width)) + 
  geom_point() + 
  geom_smooth(method = "lm")

ggplot(myris, aes(x = Sepal.Length, y = Sepal.Width)) + 
  geom_point(aes(color = Species), size = 4) + 
  theme_minimal() +
  geom_smooth(method = "lm")

ggplot(myris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) + 
  geom_point(size = 4) + 
  theme_minimal() +
  geom_smooth(method = "lm")

dev.off()

```

### More on colors
  - http://colorbrewer2.org/
  

```{r}
ggplot(myris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) + 
  geom_point(size = 4) + 
  theme_minimal() +
  geom_smooth(method = "lm") + 
  scale_color_brewer(palette = "Set1")

ggplot(myris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) + 
  geom_point(size = 4) + 
  theme_minimal() +
  geom_smooth(method = "lm") + 
  scale_color_manual(values = c("pink", "goldenrod", "steelblue"))

my.palette <- scale_color_manual(values = c("pink", "goldenrod", "steelblue"))

ggplot(myris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) + 
  geom_point(size = 4) + 
  theme_minimal() +
  geom_smooth(method = "lm") +
  my.palette

```

### Faceting

```{r}

ggplot(myris, aes(x = Sepal.Length)) + geom_histogram() + 
  facet_wrap(~ Species)

ggplot(myris, aes(x = Sepal.Length)) + geom_histogram() + 
  facet_grid(Species ~ .)

ggplot(myris, aes(x = Sepal.Length)) + geom_histogram() + 
  facet_grid(. ~ Species)

```

