---
title: "Day 7 Practice"
output: html_document
---

In this practice, let's explore data on tipping, in the `tips` data frame from the `reshape2` package.

0. First get a little familiar with the structure of the data, using `summary()`, `head()`, etc.  What are the numeric variables?  What are their units? What about factors? Are factor levels sensible?

```{r}
library(reshape2)
library(tidyr)
library(ggplot2)

mytips <- tips
summary(mytips)
head(mytips)
```

1. Create a simple scatterplot, plotting the tip amount by the total bill.
  - first plot the raw quantities
  - "save" the initial simple plot as an object
  - change the axis labels for both x and y
  - try two different methods
```{r}
scatter1 <- ggplot(mytips, aes(x = total_bill, y = tip)) + geom_point()
scatter1

scatter1 + xlab("total bill amount") + ylab("tip amount")
scatter1 + scale_x_continuous(name = "total bill amount") + scale_y_continuous(name = "tip amount")

# classic heteroskedacity
scatter2 <- ggplot(mytips, aes(x = log(total_bill), y = log(tip))) + geom_point()
scatter2 + xlab("log of total bill") + ylab("log of tip")
```

2. Continue using the previous data
  - using `scales`, plot the axes so that they follow a logarithmic pattern
  - add "break" values every 5 dollars on the x axis, and every 2 dollars on the y axis
  - finally, use `labels = dollar` (made available by the `scales` package) to format the axis numbers as dollars

```{r}
scatter1 + scale_x_log10(name = "total bill amount") + scale_y_log10(name = "tip amount")
scatter1 + scale_x_log10(name = "total bill amount", breaks = seq(0, 55, 5)) + scale_y_log10(name = "tip amount", breaks = seq(0, 10, 2))
library(scales)
scatter1 + scale_x_log10(name = "total bill amount", breaks = seq(0, 55, 5), labels = dollar) + scale_y_log10(name = "tip amount", breaks = seq(0, 10, 2), labels = dollar)

```

3. Continue to build off of the final scatterplot in the previous problem (hint: you may need to change the plot object, in order to make the following changes)
  - map gender to the color of the points
  - change the color values, using both the "brewer" and "manual" methods
  - add regression lines (regular linear regression), which also break down by gender
  - also try a loess smoother instead of regression lines
  - finally, change the overall "theme" to either the "classic", "minimal", or "bw" theme

```{r}
scatter.base <- ggplot(mytips, aes(x = total_bill, y = tip)) + scale_x_log10(name = "total bill amount", breaks = seq(0, 55, 5), labels = dollar) + scale_y_log10(name = "tip amount", breaks = seq(0, 10, 2), labels = dollar)
scatter.base

scatter.base + geom_point(aes(color = sex))
scatter.base + geom_point(aes(color = sex)) + scale_color_brewer(palette = "Set1")
scatter.base + geom_point(aes(color = sex)) + scale_color_manual(values = c("magenta", "blue"))

scatter.base + geom_point(aes(color = sex)) + geom_smooth(method = "lm", aes(color = sex)) + scale_color_manual(values = c("magenta", "blue"))

ggplot(mytips, aes(x = total_bill, y = tip, color = sex)) + scale_x_log10(name = "total bill amount", breaks = seq(0, 55, 5), labels = dollar) + scale_y_log10(name = "tip amount", breaks = seq(0, 10, 2), labels = dollar) + geom_point() + geom_smooth(method = "lm") 

scatter.base + geom_point(aes(color = sex)) + geom_smooth(method = "loess", aes(color = sex)) + scale_color_manual(values = c("magenta", "blue"))

scatter.base + geom_point(aes(color = sex)) + geom_smooth(method = "loess", aes(color = sex)) + scale_color_manual(values = c("magenta", "blue")) + theme_classic()

```


4. Now let's modify the legend for the `sex` variable, building off off the scatterplot with linear regression lines
  - using the appropriate `scale` function, change the name of the legend to "Gender"
  - change the location of the legend with the `theme()` function
  - also using the "legend.position" argument, remove the legend completely
  - finally, put the legend back to the default position, but change the levels, such that Male is given as the first level
  
```{r}
scatter.base + geom_point(aes(color = sex)) + geom_smooth(method = "lm", aes(color = sex)) + scale_color_manual(values = c("magenta", "blue"), name = "Gender") + theme_minimal() + theme(legend.position = "top") # note that order matters: put the theme() after theme_minimal()

scatter.base + geom_point(aes(color = sex)) + geom_smooth(method = "lm", aes(color = sex)) + scale_color_manual(values = c("magenta", "blue"), name = "Gender") + theme_minimal() + theme(legend.position = "none")

mytips$sex <- factor(mytips$sex, levels = c("Male", "Female"))

# changing data does not "update" scatter.base automatically
scatter.base <- ggplot(mytips, aes(x = total_bill, y = tip)) + scale_x_log10(name = "total bill amount", breaks = seq(0, 55, 5), labels = dollar) + scale_y_log10(name = "tip amount", breaks = seq(0, 10, 2), labels = dollar)
scatter.base + geom_point(aes(color = sex)) + geom_smooth(method = "lm", aes(color = sex)) + scale_color_manual(values = c("magenta", "blue"), name = "Gender") + theme_minimal()

```

5. Let's explore how this pattern might vary for different days of the week.
  - use faceting to create a separate plot for each day
  - try both the "wrap" and "grid" faceting functions
  - make a change so that the labels on the top of the facets give the full day name (e.g., "Friday", not "Fri"), and they are ordered from Thursday to Sunday
  
```{r}
scatter.base + geom_point(aes(color = sex)) + geom_smooth(method = "lm", aes(color = sex)) + scale_color_manual(values = c("magenta", "blue"), name = "Gender") + theme_minimal() + facet_wrap(~ day)

mytips$day <- factor(mytips$day, levels = c("Thur", "Fri", "Sat", "Sun"))
levels(mytips$day) <- c("Thursday", "Friday", "Saturday", "Sunday")

scatter.base <- ggplot(mytips, aes(x = total_bill, y = tip)) + scale_x_log10(name = "total bill amount", breaks = seq(0, 55, 5), labels = dollar) + scale_y_log10(name = "tip amount", breaks = seq(0, 10, 2), labels = dollar)
scatter.base + geom_point(aes(color = sex)) + geom_smooth(method = "lm", aes(color = sex)) + scale_color_manual(values = c("magenta", "blue"), name = "Gender") + theme_minimal() + facet_wrap(~ day)

# "free" scales
scatter.base + geom_point(aes(color = sex)) + geom_smooth(method = "lm", aes(color = sex)) + scale_color_manual(values = c("magenta", "blue"), name = "Gender") + theme_minimal() + facet_wrap(~ day, scales = "free_x")

scatter.base + geom_point(aes(color = sex)) + geom_smooth(method = "lm", aes(color = sex)) + scale_color_manual(values = c("magenta", "blue"), name = "Gender") + theme_minimal() + facet_wrap(~ day, scales = "free_y")

scatter.base + geom_point(aes(color = sex)) + geom_smooth(method = "lm", aes(color = sex)) + scale_color_manual(values = c("magenta", "blue"), name = "Gender") + theme_minimal() + facet_wrap(~ day, scales = "free")

scatter.base + geom_point(aes(color = sex)) + geom_smooth(method = "lm", aes(color = sex)) + scale_color_manual(values = c("magenta", "blue"), name = "Gender") + theme_minimal() + facet_grid(. ~ day)
scatter.base + geom_point(aes(color = sex)) + geom_smooth(method = "lm", aes(color = sex)) + scale_color_manual(values = c("magenta", "blue"), name = "Gender") + theme_minimal() + facet_grid(day ~ .)


```

6. Now let's look at adding information about the size of the party to the plot
  - first leave off the regression lines, to make it easier to see
  - map party size to the size of the points
  - look at both an aggregate plot and plots faceted by day
  - now add back the regression lines, but only plot a single regression line (i.e., not broken down by gender), remove the confidence interval (hint: look at the arguments for `geom_smooth()`), and make the line a dashed black line
  
```{r}
scatter.base + geom_point(aes(color = sex, size = size)) + scale_color_brewer(palette = "Set1", name = "Gender") + theme_minimal() + facet_wrap(~ day)

scatter.base + geom_point(aes(color = sex, size = size)) + geom_smooth(method = "lm", se = FALSE, color = "black", linetype = 2) + scale_color_brewer(palette = "Set1", name = "Gender") + theme_minimal() + facet_wrap(~ day)

```

