---
title: "Day 7 lecture"
output: html_document
---

# HW 6 review

# Audience for plots
  - For data exploration & understanding: yourself
  - For explanation & story-telling: others

# Today's topics
  - tweaking plots
  - using `scale_...()` functions in `ggplot2`
    - labels
    - legends
    - scales
  - adding more layers
  - annotation on plots

# Tweaking plots

## Labels
  - multiple ways to change labels (of axes, etc.)
    - change the data
    - use "convenience" functions
    - modify the scale

```{r}
library(reshape2)
library(tidyr)
library(ggplot2)
library(dplyr)

myris <- iris
myris$plant <- 1:nrow(myris)
myris <- gather(myris, myvariable, cm, 1:4)
myris <- separate(myris, myvariable, c("part", "dimension"), sep = "\\.")
myris$part <- as.factor(myris$part)
myris$dimension <- as.factor(myris$dimension)

ggplot(myris, aes(cm)) + geom_histogram()

myris$centimeters <- myris$cm

ggplot(myris, aes(centimeters)) + geom_histogram()

ggplot(myris, aes(cm)) + geom_histogram() + xlab("centimeters")

myris.wide <- select(myris, -centimeters) %>% spread(dimension, cm)

ggplot(myris.wide, aes(x = Length, y = Width)) + geom_point() + xlab("length of the part") + ylab("width of the part")

# scale_[aesthetic name]_[scale type]
ggplot(myris.wide, aes(x = Length, y = Width)) + geom_point() + 
  scale_x_continuous(name = "length of a piece of the flower")

```

## PRACTICE #1

## Other scale properties
  - labels
  - breaks
  - scaling (e.g., log)

```{r}
myris.scatter <- ggplot(myris.wide, aes(x = Length, Width)) + geom_point()
myris.scatter

myris.scatter + scale_x_continuous(labels = c("two", "four", "six", "eight"), breaks = c(2, 4, 6, 8))
myris.scatter + scale_x_continuous(breaks = 1:8, labels = c("two", "four", "six", "eight", "red", "green", "up", "down"))

myris.scatter + scale_x_continuous(breaks = seq(1, 8, .3))
myris.scatter + scale_y_continuous(breaks = seq(0, 5, .25))

myris.scatter + scale_x_log10(breaks = seq(0, 8, 2))

```

## PRACTICE #2 & #3

## Legends
  - position: using theme
  - name: using scale
  - levels: changing data

```{r}
myris.scatter2 <- ggplot(myris.wide, aes(x = Length, Width)) + geom_point(aes(color = Species), size = 4)
set1.colors <- scale_color_brewer(palette = "Set1")
myris.scatter2 
myris.scatter2 + set1.colors
myris.scatter2 + set1.colors + theme(legend.position = "top")
myris.scatter2 + set1.colors + theme(legend.position = "none")

myris.scatter2 + scale_color_brewer(palette = "Set1", name = "Iris species")

myris.wide2 <- myris.wide
levels(myris.wide2$Species)
levels(myris.wide2$Species) <- c("a", "b", "c")

ggplot(myris.wide2, aes(x = Length, Width)) + geom_point(aes(color = Species), size = 4) + set1.colors

```

## PRACTICE #4, #5, & #6

## Color control examples

```{r}
myris.scatter2 + scale_color_manual(values = c("red", "blue", "green"))

# RGB
umd.red <- rgb(224, 58, 62, maxColorValue = 255)
umd.gold <- rgb(255, 225, 79, maxColorValue = 255)
umd.black <- rgb(0, 0, 0, maxColorValue = 255)

maryland.colors <- c(umd.red, umd.gold, umd.black)

myris.scatter2 + scale_color_manual(values = c(umd.red, umd.gold, umd.black))
myris.scatter2 + scale_color_manual(values = maryland.colors)

my.colors <- c("pink", "seashell", umd.gold)
myris.scatter2 + scale_color_manual(values = my.colors)

```

## More on themes

Check out this page for LOTS of options: http://docs.ggplot2.org/current/theme.html

# Additional layers

```{r}
myris.scatter2 + geom_abline(intercept = 0, slope = 1, color = "red", linetype = 2)
myris.scatter2 + geom_hline(yintercept = 2.5, linetype = 2)
myris.scatter2 + geom_vline(xintercept = 5, linetype = 3)
myris.scatter2 + geom_abline(intercept = 0, slope = 1, color = "red", linetype = 2) + 
  geom_hline(yintercept = 2.5, linetype = 2) +
  geom_vline(xintercept = 5, linetype = 3)

```


# Annotation on plots

```{r}
myris.scatter2 + geom_hline(yintercept = 2.5, linetype = 2) + 
  annotate(geom = "text", x = 2, y = 2.7, label = "interesting threshold")

myris.scatter2 + geom_hline(yintercept = 2.5, linetype = 2) + 
  annotate(geom = "text", x = 2, y = 2.7, label = "interesting threshold") +
  annotate(geom = "rect", xmin = 5, xmax = 7, ymin = 2, ymax = 3, alpha = .5) +
  annotate(geom = "text", x = 7.5, y = 2.2, label = "<-- danger zone!", color = "red")

```

