---
title: "Day 8 Practice"
output: html_document
---

1. Write an if...else statement to test whether a string matches "right", and if so, print "you win!", but if the string matches "wrong", print "you lose!", and if it's neither, print "the result is inconclusive".

```{r}
x <- "Right"
if(x %in% c("right", "Right", "RIGHT")) {
  print("you win!")
} else if(x %in% "wrong") {
  print("you lose!")
} else {
  print("the result is inconclusive")
}



```

2. Make a copy of the iris data. Use an ifelse statement to create a new column, where:
  - if the value of Sepal.Length is less than 5, the new column should have a value of "low"
  - if the value of Sepal.Length is greater than 6, it should have a value of "high"
  - and if it's between 5 and 6 (inclusive), it should have a value of "mid"
  - make a plot in ggplot2 to confirm that you did this right
  
```{r}
myris <- iris
library(ggplot2)
myris$factor <- ifelse(myris$Sepal.Length < 5, "low", ifelse(myris$Sepal.Length > 6, "high", "mid"))

myris$factor <- ifelse(myris$Sepal.Length < 5, "low", ifelse(myris$Sepal.Length >= 5 & myris$Sepal.Length <= 6, "mid", "high"))

myris$factor <- ifelse(myris$Sepal.Length < 5, "low", ifelse(myris$Sepal.Length > 5 & myris$Sepal.Length < 6, "mid", ifelse(myris$Sepal.Length > 6, "high", NA))) %>% factor(levels = c("low", "mid", "high"))
summary(myris)

xtabs(~ Sepal.Length + factor, myris)

ggplot(myris, aes(factor, Sepal.Length)) + geom_point()
ggplot(myris, aes(factor, Sepal.Length)) + geom_boxplot()
ggplot(myris, aes(Sepal.Length)) + geom_histogram(aes(fill = factor))
```

3. Consider the iris data again.
  - make a scatterplot of Sepal.Length by Sepal.Width
  - facet the plot by Species
  - using a loop, loop through species and create three separate plots
    - in each plot, display the species as the plot title 
      (check `?labs` for how to set the plot title in a convenient way)
    - use pdf() and dev.off() so that the three plots get written out to a file
    
```{r}
myris <- iris
ggplot(myris, aes(Sepal.Width, Sepal.Length)) + geom_point() + facet_wrap(~ Species)

pdf("plots_from_loop.pdf")
#for(species in c("setosa", "virginica", "versicolor")) {
#for(species in unique(myris$Species)) {
for(species in levels(myris$Species)) {
  print(ggplot(myris[myris$Species == species, ], aes(Sepal.Width, Sepal.Length)) + geom_point() + labs(title = species))
}
dev.off()

```

4. Similar to what we did with the correlation power simulations, write code (including a loop) that creates a power simulation for comparing the means of two normally-distributed sample.
  - compare the sample with both `t.test()` and `wilcox.test()`
  - put the p-values from both tests in each simulation result
  - compare the power of the t-test vs. the power of the Wilcoxon test
  
```{r}
# set up the parameters of the simulation
n.sim <- 10000
sample.n <- 100
mean.difference <- .3
sd.x <- 1
sd.y <- 1

results <- data.frame(iteration = 1:n.sim, p.ttest = NA, p.wilcox = NA)

for(counter in 1:n.sim) {
  this.x <- rnorm(sample.n, mean = 0, sd = sd.x)
  this.y <- rnorm(sample.n, mean = mean.difference, sd = sd.y)
  this.ttest <- t.test(this.x, this.y)
  this.wilcox <- wilcox.test(this.x, this.y)
  results[counter, "p.ttest"] <- this.ttest$p.value
  results[counter, "p.wilcox"] <- this.wilcox$p.value
  if(counter %% 500 == 0) {
    print(paste("finished simulation #", counter))
  }
}
head(results)
results$ttestsig <- ifelse(results$p.ttest < .05, TRUE, FALSE)
results$wilcoxsig <- ifelse(results$p.wilcox < .05, TRUE, FALSE)
head(results)
sum(results$ttestsig)/nrow(results)
sum(results$wilcoxsig)/nrow(results)

```

