---
title: "Day 3 Practice"
output: html_document
---

1. Make a "copy" of the `iris` data set in your workspace. Use `summary()` on it, identify a factor column.

```{r}

```

2. Create different versions of the factor column, in the new data set:
   a. a numeric vector with `as.numeric()`
   b. a character vector
   
```{r}

```

3. Use `xtabs()` to see how the values in these new columns line up with the original factor column.

```{r}

```

4. Check the order of levels of Species with `levels()`. Create a new column from this factor, changing the order of the levels, using:
  a. `relevel()`
  b. `reorder()`
  c. `factor()`

```{r}

```

5. Working with missing values
   - read in the data from "CognitionPaperModifiedData.csv"
   - locate NAs
   - display the rows of the data frame where the WithinMinPairs variable has missing values
   - create a new column that's a copy of the WithinMinPairs column
   - replace all missing values in this column with zeros
   - use `xtabs()` to check this change in coding

```{r}
scottdata <- read.csv("../scottdata/CognitionPaperModifiedData.csv")
```

