---
title: "Day 3 Practice"
output: html_document
---

1. Make a "copy" of the `iris` data set in your workspace. Use `summary()` on it, identify a factor column.

```{r}
myris <- iris
summary(myris) # Species is a factor
class(myris$Species)

lapply(myris, is.factor)

is.factor(myris[ , 1])
is.factor(myris[ , 2])
is.factor(myris[ , 3])
is.factor(myris[ , 4])
is.factor(myris[ , 5])

```

2. Create different versions of the factor column, in the new data set:
   a. a numeric vector with `as.numeric()`
   b. a character vector
   
```{r}
myris$SpeciesNum <- as.numeric(myris$Species)
myris$SpeciesChar <- as.character(myris$Species)
head(myris)
summary(myris)

mysleep <- sleep
summary(mysleep)
mysleep <- mysleep[mysleep$ID !=  "4", ]
summary(mysleep)
mysleep <- droplevels(mysleep)
summary(mysleep$ID)
as.numeric(mysleep$ID)
mysleep$IDnum <- as.numeric(mysleep$ID)
xtabs(~ IDnum + ID, mysleep)
xtabs(~ as.numeric(mysleep$ID) + ID, mysleep)

```

3. Use `xtabs()` to see how the values in these new columns line up with the original factor column.

```{r}
xtabs(~ Species + SpeciesNum, myris)
xtabs(~ Species + SpeciesChar, myris)
```

4. Check the order of levels of Species with `levels()`. Create a new column from this factor, changing the order of the levels, using:
  a. `relevel()`
  b. `reorder()`
  c. `factor()`

```{r}
levels(myris$Species)
myris$SpeciesNew1 <- relevel(myris$Species, "virginica")
levels(myris$SpeciesNew1)
myris$SpeciesNumNew1 <- as.numeric(myris$SpeciesNew1)
xtabs(~ SpeciesNew1 + SpeciesNumNew1, myris)

myris$SpeciesOrder[myris$Species == "setosa"] <- 6
myris$SpeciesOrder[myris$Species == "virginica"] <- 2
myris$SpeciesOrder[myris$Species == "versicolor"] <- 4
xtabs(~ Species + SpeciesOrder, myris)
myris$SpeciesNew2 <- reorder(myris$Species, myris$SpeciesOrder)
levels(myris$SpeciesNew2)

myris$SpeciesNew3 <- factor(myris$Species, levels = c("versicolor", "virginica", "setosa"))
levels(myris$SpeciesNew3)

```

5. Working with missing values
   - read in the data from "CognitionPaperModifiedData.csv"
   - locate NAs
   - display the rows of the data frame where the WithinMinPairs variable has missing values
   - create a new column that's a copy of the WithinMinPairs column
   - replace all missing values in this column with zeros
   - try using `xtabs()` to check this change in coding (why doesn't this work?)
   - try another method for double-checking your coding

```{r}
scottdata <- read.csv("scottdata/CognitionPaperModifiedData.csv")
summary(scottdata)
scottdata[is.na(scottdata$WithinMinPairs), ]
# scottdata[!is.na(scottdata$WithinMinPairs), ] # gets rid of missing data
scottdata$CompleteWithinMinPairs <- scottdata$WithinMinPairs
identical(scottdata$CompleteWithinMinPairs, scottdata$WithinMinPairs)
scottdata$CompleteWithinMinPairs[is.na(scottdata$CompleteWithinMinPairs)] <- 0
xtabs(~ WithinMinPairs + CompleteWithinMinPairs, scottdata)
summary(scottdata)
scottdata[is.na(scottdata$WithinMinPairs), c("WithinMinPairs", "CompleteWithinMinPairs")]
```

