---
title: "Day 10 lecture/review"
output: html_document
---

# Template for working with new functions

## Some steps
  1. Find the function/package you want
    - Google
    - People you know
    - TaskViews: http://lib.stat.cmu.edu/R/CRAN/
    - Papers (contact authors if necessary)
  2. If you find a package:
    - Use the RStudio package "link"
    - Look at the CRAN page
      - Google: "CRAN <packagename>"
      - website links (URL)
      - vignettes
  3. When you find a function
    1. go to the help page (`?<functionname>`)
    2. what are the required arguments?
      - in Usage
      - no defaults!
    3. what kinds of things can go in the required arguments?
      - type of object: look in Arguments
      - format
      - `example()`
    4. try it!
    5. inspect the structure of the resulting object
    6. troubleshoot
    
## Example: factor analysis

1. Google around: decide on `factanal()`
2. Already in `stats` package
3. `?factanal`
4. Required arguments:
  - `x`
  - `factors`
  
```{r}
myris <- iris
myris.fa1 <- factanal(x = ~ Sepal.Length + Sepal.Width + Petal.Length + Petal.Width, data = myris, factors = 1)
myris.fa2 <- factanal(x = myris[, c("Sepal.Length", "Sepal.Width", "Petal.Length", "Petal.Width")], factors = 1)

myris.fa1 # same as print(myris.fa1)
myris.fa2
names(myris.fa1)
str(myris.fa1)
class(myris.fa1)

# help() : Value
```

  - reminder: 
    - ERRORS mean "I give up, there's a problem"
    - WARNINGS mean "you might not want to do that (but I did it anyway)"
  - inspect the structure of the resulting object
    - `print()` (same as just typing name of object + Enter)
    - `names()`
    - `str()`
    - look at the Value part of the help file
  - browse through optional arguments


# Template for loops (or vectorization)

Example:
  - We want a regression
  - estimate, std err, p-value
  - scatterplot with regression line
  - for each species separately

1. DO IT ONCE!
  - do whatever you want to do, on just one chunk of your data (or whatever it is that you want to iterate through)

```{r}
library(ggplot2)
myris <- iris

myris.subset <- myris[myris$Species == "setosa", ] # or filter(myris, Species == "setosa")

myris.lm <- lm(Sepal.Width ~ Sepal.Length, data = myris.subset) 
coefs <- summary(myris.lm)$coefficients
est <- coefs["Sepal.Length", "Estimate"]
std.error <- coefs["Sepal.Length", "Std. Error"]
pval <- coefs["Sepal.Length", "Pr(>|t|)"]

ggplot(myris.subset, aes(Sepal.Length, Sepal.Width)) + geom_point() + geom_smooth(method = "lm")

```

2. Figure out what MINIMALLY needs to change for each pass through a loop
  - try a copy & paste, get it to work for the next thing
  - the species
  - THAT minimal thing is what you want the counter to loop through (or index)
  
```{r}
myris <- iris

myris.subset <- myris[myris$Species == "virginica", ] # or filter(myris, Species == "virginica")

myris.lm <- lm(Sepal.Width ~ Sepal.Length, data = myris.subset) 
coefs <- summary(myris.lm)$coefficients
est <- coefs["Sepal.Length", "Estimate"]
std.error <- coefs["Sepal.Length", "Std. Error"]
pval <- coefs["Sepal.Length", "Pr(>|t|)"]

ggplot(myris.subset, aes(Sepal.Length, Sepal.Width)) + geom_point() + geom_smooth(method = "lm")

```

3. Modify to fit the loop structure
  - figure out how to get all the counter values in a vector
  - `for(COUNTER in COUNTER.VECTOR) { }`
  - set up "results storage" object
    - list
    - data.frame
    - etc.
  - how do I get the values produced in each loop?
    
```{r}
#for(this.species in c("setosa", "virginica", "versicolor")) {
#for(this.species in levels(myris$Species)) {
#for(this.species in unique(myris$Species)) {
#
all.species <- levels(myris$Species)

# storing results
# my.results <- list(est = NA, std.errors = NA, pvals = NA, plots = NA)

my.results <- data.frame(species = all.species, est = NA, std.err = NA, pval = NA)
my.plots <- list()
for(this.species in all.species) {
  myris.subset <- myris[myris$Species == this.species, ]
  myris.lm <- lm(Sepal.Width ~ Sepal.Length, data = myris.subset) 
  coefs <- summary(myris.lm)$coefficients
  
  # the values we want
  est <- coefs["Sepal.Length", "Estimate"]
  std.error <- coefs["Sepal.Length", "Std. Error"]
  pval <- coefs["Sepal.Length", "Pr(>|t|)"]
  plot <- ggplot(myris.subset, aes(Sepal.Length, Sepal.Width)) + geom_point() + geom_smooth(method = "lm") + labs(title = this.species)
  
  # stick the values in our results object
  my.results[my.results$species == this.species, c("est", "std.err", "pval")] <- c(est, std.error, pval)
  my.plots[[this.species]] <- plot
}

#this.species <- "setosa"
#  print(ggplot(myris.subset, aes(Sepal.Length, Sepal.Width)) + geom_point() + geom_smooth(method = "lm") + labs(title = this.species))

# faceting: sometimes a good alternative to looping (with plots)

ggplot(myris, aes(Sepal.Length, Sepal.Width)) + geom_point() + geom_smooth(method = "lm") + facet_wrap(~ Species)

library(grid)
library(gridExtra)
grid.arrange(my.plots[[1]], my.plots[[2]], my.plots[[3]], ncol = 1) # good for arranging multiple plots in one window/page

```

NOTE: sometimes "indices" work better than "values"

```{r}
all.species <- levels(myris$Species)

my.results <- data.frame(species = all.species, est = NA, std.err = NA, pval = NA)
my.plots <- list()

for(this.row in 1:length(all.species)) {
  myris.subset <- myris[myris$Species == all.species[this.species], ]
  myris.lm <- lm(Sepal.Width ~ Sepal.Length, data = myris.subset) 
  coefs <- summary(myris.lm)$coefficients
  
  # the values we want
  est <- coefs["Sepal.Length", "Estimate"]
  std.error <- coefs["Sepal.Length", "Std. Error"]
  pval <- coefs["Sepal.Length", "Pr(>|t|)"]
  plot <- ggplot(myris.subset, aes(Sepal.Length, Sepal.Width)) + geom_point() + geom_smooth(method = "lm") + labs(title = all.species[this.species])
  
  # stick the values in our results object
  my.results[this.row, c("est", "std.err", "pval")] <- c(est, std.error, pval)
  my.plots[[this.row]] <- plot
}

my.results
names(my.plots)
str(my.plots)
```

4. When things break: 
  - find out where it broke (if an Error)
    - last value of counter == where it broke
  - run in through single loops
    - manually set counter variable

## Vectorized version
  - same overall process
    1. DO IT ONCE
    2. Instead of finding counter variable, find arguments

```{r}

get.results <- function(chunk, DV, IV) {
  my.formula <- as.formula(paste(DV, "~", IV))
  myris.lm <- lm(my.formula, data = chunk) 
  coefs <- summary(myris.lm)$coefficients
  
  # the values we want
  est <- coefs[IV, "Estimate"]
  std.error <- coefs[IV, "Std. Error"]
  pval <- coefs[IV, "Pr(>|t|)"]
  plot <- ggplot(chunk, aes(get(IV), get(DV))) + geom_point() + geom_smooth(method = "lm") + labs(title = this.species)
  
  # stick the values in our results object
  my.results <- data.frame(est = est, std.err = std.error, pval = pval)
  #my.plots[[this.species]] <- plot
  pdf("dplyr_plots.pdf")
  print(plot)
  dev.off()
  
  return(my.results)
}

library(dplyr)

# with pipes
all.results <- myris %>% group_by(Species) %>% do(get.results(chunk = ., DV = "Petal.Width", IV = "Petal.Length")) %>% as.data.frame()

# without pipes
all.results <- as.data.frame(do(.data = group_by(myris, Species), get.results(.)))

get.results(myris)

all.results

random.thing <- function(cool.argument.name = 18) {
  sqrt(length(cool.argument.name))
}

random.thing(1:10)
random.thing()

```

- by-hand debugging

```{r}
IV <- "Petal.Width"
DV <- "Petal.Length"
chunk <- myris
  plot <- ggplot(chunk, aes(get(IV), get(DV))) + geom_point() + geom_smooth(method = "lm") + labs(title = this.species)
plot
```

# Debugging functions

```{r}
test.function <- function(chunk, DV, IV) {
  my.formula <- as.formula(paste(DV, "~", IV))
  myris.lm <- lm(my.formula, data = chunk) 
  coefs <- summary(myris.lm)$coefficients
  
  # the values we want
  est <- coefs[IV, "Estimate"]
  std.err <- coefs[IV, "Std. Error"]
  pval <- coefs[IV, "Pr(>|t|)"]
  plot <- ggplot(chunk, aes(get(IV), get(DV))) + geom_point() + geom_smooth(method = "lm") + labs(title = this.species)
  
  # stick the values in our results object
  my.results <- data.frame(est = est, std.err = std.err, pval = pval)

  return(my.results)
  
}

test.function(myris, "Sepal.Width", "Sepal.Length")

debug(test.function)

test.function(myris, "Sepal.Width", "Sepal.Length")

# capital "Q" + Enter exits debugger (aka "browser")

# can also insert browser()
test.function2 <- function(chunk, DV, IV) {
  my.formula <- as.formula(paste(DV, "~", IV))
  myris.lm <- lm(my.formula, data = chunk) 
  coefs <- summary(myris.lm)$coefficients
  
  # the values we want
  est <- coefs[IV, "Estimate"]
  std.err <- coefs[IV, "Std. Error"]
  pval <- coefs[IV, "Pr(>|t|)"]
  plot <- ggplot(chunk, aes(get(IV), get(DV))) + geom_point() + geom_smooth(method = "lm") + labs(title = this.species)
  #if(this.species == "setosa") { # if you want to selectively browse
    browser()
  #}
  # stick the values in our results object
  my.results <- data.frame(est = est, std.err = std.err, pval = pval)
  return(my.results)
}

test.function2(myris, "Sepal.Width", "Sepal.Length")

```

# Other issues

## Missing data

Best package for imputation:
  - `mice`
  - implements multiple imputation
  
# Error bars

```{r}
myris <- iris

std.err <- function(x) { sd(x)/sqrt(length(x)) }

myris.means <- myris %>% group_by(Species) %>% summarize(mean = mean(Sepal.Width), sd = sd(Sepal.Width), N = length(Sepal.Width), std.err = std.err(Sepal.Width))

myris.means$std.err2 <- apply(myris.means[, c("sd", "N")], 1, function(row) row[1]/sqrt(row[2]))

myris.means

ggplot(myris.means, aes(Species, mean)) + geom_bar(stat = "identity", aes(fill = Species)) + geom_errorbar(aes(ymin = mean - 1.96*std.err, ymax = mean + 1.96*std.err), width = .3)

ggplot(myris.means, aes(Species, mean)) + geom_pointrange(aes(ymin = mean - 1.96*std.err, ymax = mean + 1.96*std.err))

ggplot(myris.means, aes(Species, mean)) + geom_point() + geom_errorbar(aes(ymin = mean - 1.96*std.err, ymax = mean + 1.96*std.err), width = .3)

ggplot(myris.means, aes(Species, mean)) + geom_pointrange(aes(ymin = mean - 1.96*std.err, ymax = mean + 1.96*std.err)) + ylim(0, 4)

ggplot(myris.means, aes(Species, mean)) + geom_pointrange(aes(ymin = mean - 1.96*std.err, ymax = mean + 1.96*std.err)) + scale_y_continuous(limits = c(0, 4))


```

