---
title: "Day 4 Practice"
output: html_document
---

1. What are the four primary/basic data types in R?
   - Practice creating a vector of each data type
   - Using c(), concatenate pairs of these vectors together
   - What happens?  What data types are coerced to what?

```{r}
numbers <- c(1, 2, 3, 4, 5)  # or <- 1:5 or <- rep(1, 5) or <- seq(1, 5, 1)
class(numbers)
characters <- c("one", "two", "three", "four", "five")
class(characters)
factors <- as.factor(c("one", "two", "three", "four", "five"))
class(factors)
levels(factors)
factors2 <- factor(c("one", "two", "three", "four", "five"), levels = c("one", "two", "three", "four", "five"))
class(factors2)
levels(factors2)
logicals <- c(TRUE, FALSE, TRUE, TRUE, FALSE)
class(logicals)

# vectors can have only ONE data type
c(numbers, characters)
factors2
characters
c(factors2, characters)
c(as.character(factors2), characters)
c(numbers, logicals)
mean(logicals)
c(logicals, characters)
c(logicals, factors2)
c(numbers, factors2)

```

2. Make a copy of the `iris` data frame
  - Using `sample()` pull the data apart into two random batches
  - Put the pieces back together with `rbind()`
  - Put the pieces back together with `bind_rows()` from the `dplyr` package, creating an ID that keeps track of which piece is which
  
  
```{r}
nrow(iris)
rows.to.pull <- sample(1:nrow(iris), nrow(iris)/2)
iris1 <- iris[rows.to.pull, ]
iris2 <- iris[-(rows.to.pull), ] # everything *except* those row numbers in rows.to.pull

head(iris)
head(iris[, -5]) # everything *except* column 5

iris.reassembled <- rbind(iris1, iris2)
iris.reassembled <- bind_rows(iris1, iris2, .id = "batch")
head(iris.reassembled)
summary(iris.reassembled)
iris.reassembled <- bind_rows("batch1" = iris1, "batch2" = iris2, .id = "batch")
head(iris.reassembled)
```

3. Make a copy of the `iris` data frame
  - using [ , ] notation, pull the data frame into two sets of different columns
  - Put the pieces back together with `cbind()`
  - Put the pieces back together with `bind_cols()` from the `dplyr` package

```{r}
colnames(iris)
iris.left <- iris[, 1:2]
iris.right <- iris[, 3:5]

new.iris <- cbind(iris.left, iris.right)
head(new.iris)

iris.left2 <- iris[, 1:4]
iris.right2 <- iris[, 5]

new.iris2 <- cbind(iris.left2, iris.right2)
colnames(new.iris2)
colnames(new.iris2)[5]
colnames(new.iris2)[5] <- "Species"
colnames(new.iris2)

new.iris3 <- bind_cols(iris.left, iris.right)
head(new.iris3)

bind_cols(list(iris.left, iris.right))
cbind(list(iris.left, iris.right))

```

4. Make another copy of the iris data set
  - Using `data.frame()` create a (small) table that assigns a color to each species.  That is, make a column called Species and a column called Color, with one row per species
  - use `merge()` to merge the data into the copy of the iris data set
  - create another table with an additional species (and color)
  - merge this new table with iris. What happened?
  - create one more table matching species to color, with only two rows, missing setosa
  - merge this new table with the copy of iris. What happened?

```{r}
color.table <- data.frame(Species = c("setosa", "virginica", "versicolor"), Color = c("red", "blue", "violet"))

```

5. Make (yet another) copy of the iris data set
  - Create a column that takes the log of the Sepal.Width
  - Create a column that represents the multiplication of Sepal.Width by Sepal.Length
  - Create a "median split" column on Sepal.Width. In other words, calculate the median of Sepal.Width, and then create a column that indicates whether the value in that row is less than the median or not
  
  
```{r}

```
