---
title: "Day 4 Homework"
output: html_document
---

1. Find some aspect of your data that needs to be combined or merged in some way.  This could be a variety of things:
  - multiple "sheets" in an Excel
  - multiple files that should be combined into one data set
  - properties that need to be merged in to one data set. For example:
    - you have a table of subject properties (gender, handedness, etc.) that need to be merged into the item-level data
    - you have other information on countries (population, etc.) that need to be merged in
  - if you DON'T have this naturally-occurring issue, figure out a way to manufacture it. For example:
    - split up your data into chunks
    - make up some new table of information that you want to merge in
  - Explain (in words) what you are trying to do
  - Use binding/merging commands to accomplish this

```{r}
mydata <- as.data.frame(mydata[-1, ])



```

# Excursion about lists & brackets

```{r}
x <- 1:10
class(x)
x[3]

y <- as.list(1:10)
y[3]
y[[3]]
names(y) <- paste("thing", 1:10, sep = "")
names(y)
y[[3]]
y$thing3 # same as [[ ]]
y[["thing3"]] # also the same
y["thing3"] # gets the "container" as well
class(y["thing3"])
class(y[["thing3"]])
y[3]
y["thing3"]
y[c(FALSE, FALSE, TRUE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE)]
y[y == 3]

```

# Excursion
