---
title: "Day 5 Lecture"
output: html_document
---

# Outline
  - Review Homework 4
  - Merging with `merge()`
  - PRACTICE: merging
  - Manipulating and transforming variables
  - PRACTICE: transformations
  - Reshaping data & "tidy" data
  - Aggregation (e.g., cells means)
  - PRACTICE: reshaping & aggregation
  - Graphics
    - base graphics
    - ggplot2 basics
    - graphics "devices" and saving graphics
  - PRACTICE: histograms and scatterplots

# Homework 4 Review

# Merging with `merge()`

```{r}
subject.info <- data.frame(ID = 3:12, gender = c("male", "male", "female", "female", "female", "female", "male", "female", "male", "male"))
subject.info

sleep.merged <- merge(sleep, subject.info)
sleep
sleep.merged
nrow(sleep)
nrow(sleep.merged)

mysleep <- sleep
mysleep$ID <- as.numeric(mysleep$ID)

sleep.merged2 <- merge(mysleep, subject.info, all = TRUE)
sleep.merged2

```

Merging notes:
    - simplest: have ONE column in common, named the same thing (with the same values)
    - can get MUCH more complex (as needed)
      - look at "join" functions in `dplyr` for alternative functions
    - `nrow()` good check on what was gained/lost during a merge

# Data transformations    
## Creating new columns

```{r}
mysleep
mysleep$Study <- "Study1"
mysleep$noise <- rnorm(20)
mysleep
```

## Simple column transformations

```{r}
mysleep <- sleep
mysleep$extra2 <- mysleep$extra * 2
mysleep$extra3 <- log(mysleep$extra + 10)
mysleep
mysleep$extra2 <- mysleep$extra2 - 4
mysleep
mysleep$extra2 <- NULL  # deletes the column
mysleep
mysleep$extra3 <- NA
mysleep
mysleep$three <- 3
mysleep
# mysleep$four <- c(1, 2, 3) # doesn't work
mysleep$five <- c(1, 2, 3, 4) # does work (recycling)
mysleep

rep(2, 10)
rep(2:4, 10)
rep(2:4, each = 10)

rep(1:2, each = 5)

```

# Keeping "tidy" data

    - `tidyr`
    - https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html
    - http://vita.had.co.nz/papers/tidy-data.html
    - Principle: variables/data in columns, observations/cases in rows
    - wide vs. long data
      - usually long data is easier to work with
      - sometimes wide formats are useful/convenient
      - moving between the two is important!
    - `reshape2` also very powerful

## Reshaping with `tidyr` 
   
```{r}
library(tidyr)
mysleep <- sleep
mysleep
levels(mysleep$group) <- c("pre", "post")

# spread(yourdata, factor.to.spread.by, column.with.values)
mysleep.wide <- spread(mysleep, group, extra)
mysleep.wide
# QUICK PRACTICE: compute gains

# gather(yourdata, new.factor.column.name, measure.column.name, columns.to.gather)
mysleep.long <- gather(mysleep.wide, time, extra.sleep, c(pre,post))
mysleep.long
```

## PRACTICE

## Splitting up columns
  - another "tidy data" goal
  - if a column represents multiple things/datapoints, split it up!
  - use `separate()` from `tidyr`

```{r}
colnames(who)
who.long <- gather()

who.long <- separate(who.long, COL, sep = , c("colnames"))
head(who.long)

```

# Data cleaning
## A guideline template for data cleaning
  1. Get the data to read in to R
  2. Check all columns to make sure data is getting read in correctly
  3. Check NAs, decide what to do with them
  4. Remove excessive rows or other "defective" rows
  5. Reshape as needed (tend towards "long" formats)
  6. Get data out of column *names*, and into columns (variables)
  7. Fix/clean up column names
  8. Re-code factors as needed
  
## Reading data
  - find a good `read.X()` function, where X = your type of data
    - `read.table()` is the basic one
    - `read.csv()` and `read.delim()` are convenience "wrappers" for `read.table()`, for CSV and tab-delimited files, respectively
    - `read.spss()` and `read.dta()` from the `foreign` package for SPSS and Stata files
  - for messier or more primitive kinds of files in weird formats, use `readLines()`, or `scan()` to pull in the data in a very raw format, which you will then need to format in R

## Check all columns

## Check NAs
  - use summary()

## Remove rows/columns as needed

## Reshape as needed

## Get data out of column names, into variables

## Fix/improve column names

## Re-code factors as needed

# Split-apply-combine with the `dplyr` package

## Aggregation

```{r}
library(dplyr)
library(reshape2)

myff <- french_fries
head(myff)

# make long

# get "cell means"
ff.means <- summarize(group_by(myff.long, treatment, taste), mean = mean(rating))

# other summary computations

```

## Forward-pipe operator (from `magrittr` package)
["This is not a pipe"](https://en.wikipedia.org/wiki/The_Treachery_of_Images)

  - `%>%` operator
  - effect: pass the results of the thing on the left to the first argument of the thing on the right
  - very commonly used in `dplyr`

```{r}
x <- 1:10
mean(log(x + 2))

(x + 2) %>% log() %>% mean()

summarize(group_by(myff.long, treatment, taste), mean.rating = mean(rating, na.rm = TRUE))

# same as above, using the %>% operator
group_by(myff.long, treatment, taste) %>% summarize(mean.rating = mean(rating, na.rm = TRUE))

```

## Subsetting with `dplyr`
  - `filter()` is for subsetting rows
  - `select()` is for subsetting columns

```{r}
sleep
sleep[sleep$group == 1, ]
filter(sleep, group == 1) # same results

sleep[sleep$group == 1 & sleep$extra > 0.2, ]
filter(sleep, group == 1, extra > 0.2) # same results

sleep[ , c("ID", "extra")]
select(sleep, ID, extra)
select(sleep, 3, 1)
select(sleep, c(3, 1))
select(sleep, c("ID", "extra")) # doesn't work
select(sleep, c(ID, extra))
select_(sleep, "ID", "extra") # "underscore" variants of dplyr functions:
                              #   - can use strings to identify objects
                              #   - useful for more programmatic applications

```


# Notes on complex objects

  - use `names()` and `str()` to look at the structure of a complex object
  - most complex objects have a list-like structure
  - S3 vs. S4 classes
    - `$` notation works for S3
    - Try `@` instead for S4 classes
    - For more on S4: http://adv-r.had.co.nz/OO-essentials.html

```{r}
myff <- french_fries
results <- lm(potato ~ treatment, data = myff)
results
summary(results)
# plot(results)
class(results)
names(results)
str(results)
plot(results$residuals)
results.summary <- summary(results)
names(results.summary)
results.summary$adj.r.squared
results.summary$coefficients
class(results.summary$coefficients)
condition.pvals <- results.summary$coefficients[c("treatment2", "treatment3"), "Pr(>|t|)"]
condition.pvals

# str(summary(results))
# ?lm
```

The adjusted $R^2$ for this model is `r results.summary$adj.r.squared`.

# Graphics overview

http://accidental-art.tumblr.com/

## Two systems

  - "base" graphics
  - "grid" graphics
  - See *R Graphics* by Paul Murrell for a very thorough "deep-dive" into these implementations:  https://www.stat.auckland.ac.nz/~paul/RG2e/

### Base graphics
  - more "primitive"
  - can also be much faster than grid
  - lots of isolated plot functions
  - lots of weird little parameters to worry about
  - some handy quick-plot functions from base graphics:
    - `plot()` : a generic function that will often produce something interesting, can do scatterplots with a formula `y ~ x`
    - `hist()` : histogram of a vector
    - `boxplot()` : boxplot, using a formula `variable ~ grouping.factor`

### Grid graphics
  - better for more "high-level" plots
  - much more complex underlying system (IMHO)
  - more powerful and extendable
  - the basis of extremely popular packages such as `lattice` and `ggplot2`

## Graphics packages

  - many packages will have some kind of graphics functions in them, for visualizing specific kinds of things
  - these will often be in base graphics, "pre-canned", etc.
  - the two best more general-purpose graphics packages are `lattice` and `ggplot2`
  - `ggplot2` is another Hadley Wickham project, and has the most documentation, examples, and is currently most widely used

# Using `ggplot2`

## A "Grammar of Graphics"
  - based on work by Leland Wilkinson, a "theory" of graphic display of data
  - Hadley implemented this framework as his dissertation
  - `ggplot2` is the overhauled version of the original implementation, and it is still under active development
  - the idea is to implement a general system for mapping data variables to *aesthetics* (Wilkinson's term), i.e., visual dimensions (x, y, color, etc.)
  - allows for high-level control of the structure of a plot, to allow for flexible, powerful visualizations
  
## Basic structure and concepts

  - unlike base graphics, the *value* of `ggplot2` objects is a complex object, and `print()`-ing them renders the graph.
  - the value of the object is built up by adding *layers*
  - start with a call to `ggplot()` that specifies the data and some mappings of variables to aesthetics
  - add "geoms" to actually render something
  - add other things for fine control over graphical parameters, etc.

## Graphing the `iris` package

```{r}
library(ggplot2)
myris <- iris
summary(myris)

ggplot(myris, aes(x = Sepal.Length)) + geom_histogram()

ggplot(myris, aes(x = Sepal.Length)) + geom_histogram(binwidth = .5)

ggplot(myris, aes(x = Sepal.Length)) + geom_histogram(binwidth = .01)

mybaseplot <- ggplot(myris, aes(x = Sepal.Length))

mybaseplot + geom_histogram(binwidth = .2)

names(mybaseplot)
str(mybaseplot)

base.density <- mybaseplot + geom_density(adjust = .5, fill = "grey")

mydensity <- geom_density(adjust = .5, fill = "grey")

mybaseplot + mydensity

```

```{r}
colors()
```

```{r}
pdf("bunch of plots.pdf")
ggplot(myris, aes(x = Sepal.Length, y = Sepal.Width)) + geom_point()

ggplot(myris, aes(x = Sepal.Length, y = Sepal.Width)) + geom_point(color = "papayawhip") + theme_classic()

ggplot(myris, aes(x = Sepal.Length, y = Sepal.Width)) + 
  geom_point(aes(color = Species), size = 10) + 
  theme_minimal()

ggplot(myris, aes(x = Sepal.Length, y = Sepal.Width)) + 
  geom_point(aes(color = Species, size = Petal.Width)) + 
  theme_minimal()

ggplot(myris, aes(x = Sepal.Length, y = Sepal.Width)) + 
  geom_point() + 
  geom_smooth()

ggplot(myris, aes(x = Sepal.Length, y = Sepal.Width)) + 
  geom_point() + 
  geom_smooth(method = "lm")

ggplot(myris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) + 
  geom_point(size = 4) + 
  theme_minimal() +
  geom_smooth(method = "lm")
dev.off()

```

### Saving plots

```{r}
pdf("myplot.pdf", width = 10, height = 4)
ggplot(myris, aes(x = Sepal.Length)) + geom_histogram() + 
  facet_grid(Species ~ .)
dev.off()

?Devices
```

### Faceting

```{r}

ggplot(myris, aes(x = Sepal.Length)) + geom_histogram() + 
  facet_wrap(~ Species)

ggplot(myris, aes(x = Sepal.Length)) + geom_histogram() + 
  facet_grid(Species ~ .)

ggplot(myris, aes(x = Sepal.Length)) + geom_histogram() + 
  facet_grid(Species ~ .)

```

